Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-07-28 ms.tex
More DM related text

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE) gamma-ray emission. Observations of radio halos and relics in clusters indicate populations of relativistic electrons, which are either injected into the intra-cluster medium and accelerated by turbulent shocks or secondaries from interactions of cosmic-ray hadrons (protons and ions) with the intra-cluster gas. Those interactions would also produce gamma rays through the decay of neutral pions. Reported here are observations of the Coma cluster of galaxies with the VERITAS array of imaging Cherenkov telescopes. The observations were made between March and May 2008. Radio observations at 1.41 GHz taken with the Green Bank Telescope and X-ray observations with the ROSAT All-Sky survey was used to a priori select regions of interest in the field of view. No significant VHE gamma-ray emission from the Coma cluster was detected by VERITAS. Several emission regions are considered, including the cluster core as both unresolved (point-like) and spatially extended, and a select few cluster members. The flux upper limits at the 99\% confidence level measured are on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$, which is X times higher than the model-predicted minimum gamma-ray flux.

The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray particles and magnetic fields in the cluster. The relative cosmic-ray pressure $X_{\mathrm{CR}}$ is less than\ldots. The observations imply a lower limit on the intra-cluster magnetic field on the order of a $\mu$G depending on the magnetic decline and gamma-ray spectral index. Finally, a constraint on the thermally averaged product of the total self-annihilation cross section and the relative velocity of the dark-matter particles is derived, $\expval{\sigma v}\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}>X$ GeV/c$^{2}$.
\end{abstract}

@@

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of the cluster mass is in the form of dark matter. Recent works \citep{article:EvansFerrerSarkar:2004, article:BergstromHooper:2006} have shown that the gamma-ray emissivity from DM annihilations is expected to be proportional to the DM density along the line of sight, making clusters of galaxies potentially good candidates for DM searches, despite their large distances.

While several observations of clusters of galaxies have been made with satellite-borne and ground-based telescopes, a detection of gamma-ray emission from a cluster has yet to be made.  Observations with EGRET \citep{article:Sreekumar_etal:1996, article:Reimer_etal:2003} and the Fermi Gamma-ray Space Telescope \citep{article:Ackermann_etal:2010} have provided upper limits on the gamma-ray fluxes (typically $\sim10^{-9}$ ph cm$^{2}$ s$^{-1}$ for Fermi observations) for several galaxy clusters in the MeV to GeV band. Upper limits on the VHE gamma-ray flux from a few select clusters, including the Coma cluster, have been provided by observations with ground-based imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Perkins_etal:2006, inproc:Perkins_etal:2008, article:Aharonian_etal:2009, article:Aleksic_etal:2010}.

@@

\section{Emission from Dark Matter Annihilations}

>>>>

\begin{abstract}
Theoretical models of galaxy clusters suggest they may be sources of very high-energy (VHE) gamma-ray emission. Observations of radio halos and relics in clusters indicate populations of relativistic electrons, which are either injected into the intra-cluster medium and accelerated by turbulent shocks or secondaries from interactions of cosmic-ray hadrons (protons and ions) with the intra-cluster gas. Those interactions would also produce gamma rays through the decay of neutral pions. Reported here are observations of the Coma cluster of galaxies with the VERITAS array of imaging Cherenkov telescopes. The observations were made between March and May 2008. Radio observations at 1.41 GHz taken with the Green Bank Telescope and X-ray observations with the ROSAT All-Sky survey was used to a priori select regions of interest in the field of view. No significant VHE gamma-ray emission from the Coma cluster was detected by VERITAS. Several emission regions are considered, including the cluster core as both unresolved (point-like) and spatially extended, and a select few cluster members. The flux upper limits at the 99\% confidence level measured are on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$, which is two orders of magnitude higher than model-predicted minimum gamma-ray fluxes.

The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray particles and magnetic fields and in the Coma cluster of galaxies. The relative cosmic-ray pressure, $X_{\mathrm{CR}}=\expval{P_{\CR}}/\expval{P_{\rmn{th}}}$, is \ldots. The observations imply a lower limit on the intra-cluster magnetic field on the order of a $\mu$G, depending on the magnetic decline and gamma-ray spectral index. Finally, a constraint on the thermally averaged product of the total self-annihilation cross section and the relative velocity of the dark-matter particles, $\expval{\sigma v}$, is derived and found to be $\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$ GeV/c$^{2}$.
\end{abstract}

@@

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of the cluster mass is in the form of dark matter. Recent works \citep{article:EvansFerrerSarkar:2004, article:BergstromHooper:2006} have shown that the gamma-ray emissivity from DM annihilations is expected to be proportional to the DM density along the line of sight, making clusters of galaxies potentially good candidates for DM searches. This despite their large distances compared to other common targets for DM searches, such as dwarf spheriodal galaxies \citep{article:Strigari_etal:2007, article:Acciari_etal:2010, article:Aliu_etal:2009} or the Galactic center \citep{article:Kosack_etal:2004, article:Aharonian_etal:2006, article:Aharonian_etal:2009b}.

While several observations of clusters of galaxies have been made with satellite-borne and ground-based telescopes, a detection of gamma-ray emission from a cluster has yet to be made.  Observations with EGRET \citep{article:Sreekumar_etal:1996, article:Reimer_etal:2003} and the Fermi Gamma-ray Space Telescope \citep{article:Ackermann_etal:2010} have provided upper limits on the gamma-ray fluxes (typically $\sim10^{-9}$ ph cm$^{2}$ s$^{-1}$ for Fermi observations) for several galaxy clusters in the MeV to GeV band. Upper limits on the VHE gamma-ray flux from a few select clusters, including the Coma cluster, have been provided by observations with ground-based imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Perkins_etal:2006, inproc:Perkins_etal:2008, article:Aharonian_etal:2009a, article:Aleksic_etal:2010}.

@@

\section{Emission from Dark Matter Annihilations}
When the size of the integration region is increased, the limits on $\expval{\sigma v}$ result from a competition between the gain in the astrophysical factor $\expval{J}$ and the integrated background. For integration regions larger than 0.2$^{\circ}$ in radius, the astrophysical factor no longer compensate for the increased count of background events and the signal-to-noise deteriorates.
